Spatial variation of the native colon microbiota in healthy adults

### 2017-10-30 submission/manuscript.Rmd
10 31 pat edits results

#### Microbial membership and diversity of the proximal and distal colon 
Lumenal and mucosal samples were collected from the proximal and distal colon of 20 healthy humans that had not undergone bowel preparation (Figure 1). Participants also collected stool at home one week prior to the procedure. To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample. As expected, each site was primarily dominated by _Firmicutes_ and _Bacteriodetes_ (Figure 2A) @LloydPrice2016. Samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B). For example, the proximal mucosa was more diverse than the distal for some individuals while the opposite was true for others. Therefore we could not identify a clear pattern of changes in microbial diversity along the gut axis.

To compare similarity between sides (proximal or distal) or sites (lumen or mucosa), we calculated distances from Operational Taxonomic Unit (OTU) abundances and compared these distances for all individuals. Again, across all patient samples we observed a range of distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis. However, when comparing median distances between the proximal lumen and mucosa, the proximal versus distal lumen, the proximal versus distal mucosa, and the distal lumen and mucosa, we found that the proximal lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment).

#### Fecal samples resemble lumenal samples from the distal colon 
Next, we calculated distances to examine how each sample compared to the home-collected feces. Amidst variability between patients, we did identify significantly smaller distance between the distal lumenal sample and the feces (Figure 3B, P < 0.05, Wilcoxon, BH adjustment). Furthermore, there was an even larger difference in the comparisons of the distal mucosa to the feces, indicating that the mucosa is different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment). To determine what factors may be driving the differences seen among the samples, we compared distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal). We found that samples from one individual were far more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon [@Eckburg2005; @deCarcer2010; @Zhang2013]. Thus interpersonal variation between subjects drove the differences between samples more than sample site or location. Overall, the results comparing the structure of the communities suggest that the contents of the distal lumen are most representative of the patient's feces, and the microbes remaining on the mucosa are more distinct.

#### Random Forest classification models identify important Operational Taxonomic Units (OTUs) on each side
To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest models trained using OTU abundances. We used 10-fold cross validation to build the first model to classify the lumenal versus mucosal samples for the proximal and distal sides, independently (Figure S1A). The models performed well when classifying these samples (proximal AUC = 0.764, distal AUC = 0.908). The OTUs that were most predictive of each site were identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the proximal lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4A). The model classifying the distal lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera (Figure 4B). These results indicated that there were fine differences between the different sites of the colon, and that these could be traced to specific OTUs on each side.

Next, we built a model to differentiate the proximal and distal lumenal samples using 10-fold cross validation. The model performed best when distinguishing the proximal versus distal mucosa (Figure S1B, AUC = 0.850) compared to the proximal versus distal lumen (AUC = 0.580). OTUs that were differentially abundant between the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5A). Differentially abundant OTUs of the proximal and distal lumen included three OTUs of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5B). This analysis found that some of the same OTUs that are distinct between the mucosa and lumen also helped to differentiate between the two sides- such as _Anaerococcus_ and _Finegoldia_.

>>>>

#### Microbial membership and diversity of the proximal and distal colon 
Luminal and mucosal samples were collected from the proximal and distal colon of 20 healthy humans that had not undergone bowel preparation (Figure 1). Subjects also collected stool at home one week prior to the procedure. To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on DNA extracted from each sample. As expected, each site was primarily dominated by _Firmicutes_ and _Bacteriodetes_ (Figure 2A) @LloydPrice2016. Samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B). For example, the proximal mucosa was more diverse than the distal for some individuals while the opposite was true for others. Therefore we could not identify a clear pattern of changes in microbial diversity along the gut axis.

To compare similarity between the proximal and distal sides and within the lumen and mucosa, we compared the community structure of these sites based on the relative abundances of individual Operational Taxonomic Units (OTUs). Across all subjects we observed wide variation when comparing sample locations (Figure 3A). Those ranges did not follow a clear pattern on an individual basis. However, when comparing median dissimilarity between the communities found in the proximal lumen and mucosa, the proximal and distal lumen, the proximal and distal mucosa, and the distal lumen and mucosa, we found that the proximal lumen and mucosa were most similar to each other than to the other samples (P < 0.005, Wilcoxon, BH adjustment).

#### Fecal samples resemble luminal samples from the distal colon 
Next, we compared the luminal and mucosal samples to the fecal sample of each subject. Amidst the large inter-subject variation, we did identify significantly less dissimilarity between the distal luminal sample and the feces (Figure 3B, P < 0.05, Wilcoxon, BH adjustment). Furthermore, there was an even larger difference in the communities found in the distal mucosa compared to the fecal communities, indicating that the mucosa is as different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment). These results suggest that the contents of the distal lumen were most representative of the subjects' feces, and the mucosal microbiota are distinct from the fecal and luminal communities.

#### Interpersonal community variation is greater than the variation between sites
To determine what factors may have driven the differences seen among the samples, we compared the community dissimilarity between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal). We found that samples from one individual were far more similar to each other than to matched samples from the other subjects (Figure 3C); this is consistent with previous human microbiome studies that have sampled multiple sites of the human colon [@Eckburg2005; @deCarcer2010; @Zhang2013]. Thus interpersonal variation drove the differences between samples more than whether the sample came from the proximal or distal side of the colon or from the lumen or mucosa.

#### Random Forest classification models identify important Operational Taxonomic Units (OTUs) on each side
To identify OTUs that were distinct at each site, we constructed several Random Forest models trained using OTU relative abundances. We used 10-fold cross validation to build the first model to classify the luminal versus mucosal samples for the proximal and distal sides, independently (Figure S1A). The models performed well when classifying these samples (proximal AUC = 0.764, distal AUC = 0.908). The OTUs that were most predictive of each site were identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the proximal lumen and mucosa, OTUs affiliated with the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and _Enterobacteraceae_ families were included in the best model (Figure 4A). The model to differentiate between the distal lumen and mucosa included OTUs affiliated with the _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and _Anaerococcus_ families (Figure 4B). These results indicated that there were fine differences between the different sites of the colon, and that these could be traced to specific OTUs on each side.

Next, we built a Random Forest model to differentiate the proximal and distal lumenal samples using 10-fold cross validation. The model performed best when distinguishing the proximal versus distal mucosa (Figure S1B, AUC = 0.850) compared to the proximal versus distal lumen (AUC = 0.580). OTUs that were differentially abundant between the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5A). Differentially abundant OTUs of the proximal and distal lumen included three OTUs of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5B). This analysis found that some of the same OTUs that are distinct between the mucosa and lumen also helped to differentiate between the two sides- such as _Anaerococcus_ and _Finegoldia_.
